---
title: "Exploring Gun Violence Trends in the United States"
author: "Yuelin Long (yl3181), Marisa Sobel (ms5533), Eileen Shea (eas2297), Zanis Fang (zf2213), Devon Morgan (dm3175)"
date: "12/3/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(httr)
library(plotly)
library(readxl)
library(patchwork)
library(viridis)

```

## Motivation
Gun violence has been of increasing concern across the United States over the past decade. Our objective was to develop an online resource to explore general trends in gun violence and identify any important variables associated with gun violence incidence. The target audience for this webpage includes the general public, gun violence activists and legislators. The goal of this resource is to raise awareness on the widespread impact of gun violence, and inform future public health interventions and policy changes. 

## Background
+ The [American Public Health Association (APHA)](https://www.apha.org/topics-and-issues/gun-violence) states that gun violence "is a leading cause of premature death in the U.S." More than 38,000 individuals are killed by guns each year, and another 85,000 are injured.
+ Already in 2018, the [Gun Violence Archive](https://www.gunviolencearchive.org) has reported 52,422 gun violence incidents in the United States; 325 of these incidents were mass shootings.
+ Since 1996, the [Dickey Amendment](https://en.wikipedia.org/wiki/Dickey_Amendment) has acted as a de facto ban on gun violence research by the [Centers for Disease Control and Prevention (CDC)](https://www.cdc.gov). 

## Related Work
Gun violence has received widespread media attention in recent years, particularly due to the increased occurrence of mass shootings. Preventing firearm injury and death has increasingly been considered a major public health issue.

Inspiration was drawn from a variety of background sources including:

*  [Gun Violence Archive](https://www.gunviolencearchive.org/), a nonprofit focused on providing data on gun related incidents in the U.S., which provides helpful visuals related to gun violence incidents across the country.
*  [Wintemute, G. J. (2015). The epidemiology of firearm violence in the twenty-first century United States. Annual Review of Public Health, 36, 5-19.](https://www.annualreviews.org/doi/full/10.1146/annurev-publhealth-031914-122535)
*  [Gun Violence - National Institute for Justice](https://www.nij.gov/topics/crime/gun-violence/pages/welcome.aspx)
*  [Giffords Law Center](https://lawcenter.giffords.org)


The Data Science methods used for this project draw from topics covered throughout P8105. In particular, inspiration on some of the map visualizations was drawn from the in-class visualizations on the Airbnb dataset. Further, as a team comprised of biostatisticians, epidemiologists and environmental health scientists in training, we successfully drew from different areas of expertise to inform the direction of the project. 

## Initial Questions
Questions to address in the project include:

*  How do rates of gun violence differ across states? Which areas have the highest concentrations of gun violence? 
*  How have gun violence rates changed over time? 
*  Which age groups and races are most impacted by gun violence mortality, and has this changed over time? 
*  How does gun law strength relate to firearm mortality? How do gun law strengths differ across the U.S.?
*  How are unemployment rates related to firearm mortality? 
*  How are gun license approval rates related to firearm mortality and law strength? 
*  How does the gun application percentage(proportion of people applying for the license in total population) relate to firearm mortality and law strength? 

## Data Sources and Cleaning Methods
Data from five sources were considered for this project. The data sources, scraping methods and cleaning procedures for each dataset are described below. 

*  The Github repository for this project can be found [here](https://github.com/ChristineLong/p8105_Final_Project). 

*  The final website can be found [here](https://mhsobel.github.io/p8105_Final.github.io/), and corresponding Github repository for the website is located [here](https://github.com/mhsobel/p8105_Final.github.io).

*  The final report is housed in a repository [here](https://github.com/devon-m2/p8105_Gun-Violence_Final-Report). 

### Data Download
All datasets can be downloaded from the following Google Drive [link](https://drive.google.com/drive/folders/12ccHT2Xsxf_5Gwm8pL4CarYoq6PlLoTQ?usp=sharing).

Some datasets pull from active URLs, so corresponding data is not housed in the Google Drive folder. Corresponding code should pull this data.  

To re-run the report, download the corresponding data into a folder named "data" also housed in the R project.

### Dataset 1: Gun Violence Incident Data (Gun Violence Archive) 
Data on specific Gun Violence incidents throughout the U.S. was downloaded from a Kaggle data repository [here](https://www.kaggle.com/jameslko/gun-violence-data), which pulls data from the [Gun Violence Archive](http://www.shootingtracker.com/). Gun Violence Archive is a non-profit formed in 2013 to provide free online public access to accurate information about gun-related violence in the United States. The archive collects information on gun violence incidents "from over 2,500 media, law enforcement, government and commercial sources daily in an effort to provide  near-real time data about the results of gun violence. GVA in an independent data collection and research group with no affiliation with any advocacy organization." The dataset accessed from Kaggle scraped and partially tided the data from Gun Violence Archive. 

#### Cleaning
Data cleaning steps included making individual variables for day, month, and year and creating new variables for the number affected in a  given gun violence incident (`n_affected`) and that classify mass shooting status (`mass_shooting`). 

```{r message = FALSE}
gun_violence_data = read_csv("./data/gun_violence_data_2013_2018.csv")

gun_v_tidy = 
	gun_violence_data %>% 
	select(date:city_or_county, n_killed, n_injured, latitude, longitude) %>% 
	separate(date, into = c("year", "month", "day"), sep = "-") %>% 
	mutate(n_affected = n_killed + n_injured) %>%
	filter(n_affected > 0) %>% 
	mutate(mass_shooting = ifelse(n_affected >= 4, "Yes", "No"))
```


### Dataset 2: CDC Firearm Mortality Data 
Data on Firearm Mortality was obtained from the [CDC Wonder data query](https://wonder.cdc.gov/ucd-icd10.html). Data was queried on November 13, 2018. Three queries were carried out including the following variables: 

1. Firearm mortality: All - `state`, `year`, `n_deaths` (number of firearm deaths), `n_population` (state population), and `crude_rate` (provided by CDC - crude_rate = [n_deaths/n_population]*100000,
2. Firearm mortality: By Age groups and Race - `state`, `year`, `n_deaths` (number of firearm deaths), `n_population` (state population), `crude_rate`, `age_group` (age ranges for victims), `race` (race of victim - Black or African American, White, American Indian or Alaska Native, Asian or Pacific Islander), and 
3. Firearm mortality: By Age groups and Hispanic origin - `state`, `year`, `n_deaths` (number of firearm deaths), `n_population` (state population), `crude_rate`, `age_group` (age ranges for victims), and `hispanic_origin` ("Not hispanic or latino", "Unreliable", "Hispanic or latino"). 

Note that these data were pulled in three batches because the CDC Wonder database puts limits on the number of variables to query at once.

Centers for Disease Control and Prevention, National Center for Health Statistics. Underlying Cause of Death 1999-2016 on CDC WONDER Online Database, released December, 2017. Data are from the Multiple Cause of Death Files, 1999-2016, as compiled from data provided by the 57 vital statistics jurisdictions through the Vital Statistics Cooperative Program.

#### Cleaning
The following steps were used to clean the CDC Firearm data, including adding state abbreviations, and changing `age_groups` into a factor variable. 
```{r message = FALSE}
# Detailed Firearm Mortality Datatsets by Race, Age
firearm_mortality = read_csv("./data/cdc_firearm_mortality_data.csv", na = "Unreliable") %>% 
    janitor::clean_names() %>% 
    select(-ten_year_age_groups_code, -injury_mechanism_all_other_leading_causes_code, -race_code, death_cause= injury_mechanism_all_other_leading_causes) %>% 
    mutate(ten_year_age_groups = factor(ten_year_age_groups, levels = c("1-4 years", "5-14 years", "15-24 years", "25-34 years", "35-44 years", "45-54 years", "55-64 years", "65-74 years", "75-84 years", "85+ years")))

# Summary Firearm Mortality Dataset by Year
firearm_mortality_summary = read_excel("./data/cdc_firearm_all_ages.xlsx") %>% 
    janitor::clean_names() %>% 
    select(-year_code, -injury_mechanism_all_other_leading_causes_code, death_cause= injury_mechanism_all_other_leading_causes)

# Created table with state name and state abbreviations crosswalk
st_crosswalk = tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
     bind_rows(tibble(state = "District of Columbia", abb = "DC"))

# Joined abbreviation dataset with summary firearm mortality dataset
clean_firearm_mortality = left_join(firearm_mortality_summary, st_crosswalk, by = "state") %>% 
    rename(state_abb = abb)
```

### Dataset 3: Gun Law Strength and State Scoring
Gun law scores of states data was taken from the [Giffords Law Center](https://lawcenter.giffords.org/scorecard/#rankings). Gun law scores were developed using "a comprehensive grading rubric that assigns positive point values to gun safety policies, such as private-sale background checks and extreme risk protection orders, and negative point values to dangerous laws, such as permitless concealed carry". 

The search for gun law strength came from an interest in the state-specific gun climate. The Gifford’s Law Center Gun Law Scorecard resulted from a quick Google search, and neatly implemented many of the factors we had thought independly would influence the gun climate. The Scorecard factors in different types of gun policies that protect citizens from gun violence, and others that protect tangentially, such as domestic violence.

#### Cleaning
The Scorecard report exists on the Gifford’s Law Center’s website as a table. A css table was extracted from the URL that included all the data from the Scorecard in an already tidied manner, as each row was a different state and each column was a different variable. Variables were cleaned up and renamed for ease. For sake of easier comparisons, the grade scale was collapsed from 10 levels to 5, removing the pluses and minuses and keeping the alphabetic grade. The death rate and death rate rank were removed as we were obtaining this information from other sources. 

```{r message = FALSE}
# website URL
url = "https://lawcenter.giffords.org/scorecard/#rankings"
gun_climate_url = read_html(url)

# extract table and clean data
gun_climate_data = 
	gun_climate_url %>% 
  html_nodes(css = "table") %>% 
  .[[1]] %>% 
  html_table(header = TRUE) %>% 
  as.tibble() %>% 
	janitor::clean_names() %>% 
	rename(
		law_strength = gun_law_strength_ranked, 
		grade_2017 = x2017grade, 
		death_rate_rank = gun_death_rate_ranked, 
		death_rate = gun_death_rate_per_100k) %>% 
	mutate(
    grade_2017 = factor(grade_2017, 
                      levels = c("A", "A-", "B+", "B", "C+", "C", "C-", "D", "D-", "F")), 
    grade_2017 = fct_collapse(grade_2017, 
    													A = c("A", "A-"), 
    													B = c("B+", "B"), 
    													C = c("C+", "C", "C-"), 
    													D = c("D", "D-")), 
    state_abb = state.abb)
```


### Dataset 4: Gun Approval Rate and Licensing of States
Data on gun license applications was taken from the [The NICS background checks](https://www.statista.com/statistics/249687/number-of-background-checks-done-by-the-nics-in-the-us-by-state/) source, used by Federal Firearms Licensees (FFLs) to instantly determine whether a prospective buyer is eligible to buy firearms or explosives. This data is used to approximate how many people tried to apply for license in each state. 

Data on [The number of federal firearms licensees in the U.S.](https://www.statista.com/statistics/215670/number-of-federal-firearms-licensees-in-the-us-by-state/) was downloaded from Statistica.com. This source pulls data from the U.S. Bureau of Alcohol, Tobacco, Firearms and Explosives (ATF), and is used to approximate how many people actually got approved for gun licenses in each state. 

Data on total population was taken from [The total population across different states in the U.S.](https://www.statista.com/statistics/183497/population-in-the-federal-states-of-the-us/) source. It was used to standardize number of approved license and application numbers in different states. 

#### Cleaning
The following steps were used to clean the gun approval rate and license data, including renaming `state` variable. The approval rate uses the proportion of licensees in the total population of the state to indicate the difficulties in getting apprroved for gun license. The application rate uses the proportion of background checks in the total population of the state to indicate people's willingness in applying for guns. For consistency purposes, District of Columbia was not included.

```{r message = FALSE}
# Create a function to import and clean data
clean_fun = function(address, area){
  readxl::read_xlsx(address, sheet = "Data", range = area) %>% 
  rename(state = X__1) %>% 
  janitor::clean_names()
}

# Import the total population data 
total_pop = clean_fun("./data/population-in-the-states-of-the-us-as-of-2017.xlsx", "B5:C56")

# Import the approved license data
approved_lic = clean_fun("./data/number-of-federal-firearms-licensees-in-the-us-in-2017-by-state.xlsx", "B5:C57") %>% 
  filter(state != "Other Territories")

# Import the background check data
back_check = clean_fun("./data/nics-background-checks-done-by-us-firearms-licensees-2017-by-state.xlsx", "B5:C56")

# Merge the datasets together
gun_lic = inner_join(approved_lic, back_check, by = "state") %>% 
  inner_join(total_pop, by = "state") %>% 
  mutate(approval_rate = number_of_federal_firearms_licensees/number_of_residents_in_millions/1000000,
         application_rate = number_of_background_checks/number_of_residents_in_millions/1000000)
```


### Dataset 5: Unemployment data


#### Cleaning


### Dataset 6: U.S. Chronic Disease Indicators (CDI)


#### Cleaning



## Exploratory Analyses
Exploratory analyses included summarizing the number of gun violence incidents by states and visualizing this through a bubble map, as well as looking at time trends. A decision was ultimately made to favor the interactive map over the bubble map because of its greater descriptive power and to not include time trend analyses regarding this dataset since the span from 2014-2017 (only full years of data collection) proved too short to notice anything significant/interesting.

Decided to use the data on firearm mortality from the CDC. 

Decided to not focus on subtypes of gun violence, including mass shootings versus domestic violence, but gun violence in general. We acknowledge that risk factors could differ across sub-types. Important to consider in the future -- 
overall 

Preliminary maps were made with ggmap showing a gradient of gun law strength as well as the categorical grades. 

 
##### Numerical variables
```{r }
gun_control = gun_climate_data %>% 
  inner_join(gun_lic, by = "state") %>% 
  inner_join(clean_firearm_mortality, by = "state")

skimr::skim(gun_control) %>%
            select(variable:stat,value) %>%
            filter((stat != "hist") , (stat != "top_counts"),(type %in% c("numeric","integer"))) %>% 
            spread(key = stat,value = value) %>% 
  knitr::kable(digits = 1) 
```

##### Categorical variables
```{r }
skimr::skim(gun_control) %>%
            select(variable:stat,value) %>%
            filter((stat != "hist") , (stat != "top_counts"),(type %in% c("character","factor"))) %>% 
            spread(key = stat,value = value) %>% 
  knitr::kable(digits = 1) 
```

As we can see, the dataset has no missing data. All the information, including population statistics, background check / applications statistics, approved licensees statistics,  are generalized for different states (except for D.C.).

## Regression Analyses 
Section 4: Regression Analyses


## Discussion and Results
Results from our analyses are organized in four sections, housed on separate pages of the website: 

+ Section 1: Gun Violence - Overview
+ Section 2: Firearms Mortality
+ Section 3: Regression Analysis
+ Section 4: Gun Control by State

### Section 1: Gun Violence - Overview
Section 1 provides an overview of Gun Violence in the U.S., drawing from **Dataset 1: Gun Violence Incident Data (Gun Violence Archive)**. The primary goal of this section is to highlight the widespread nature of gun violence throughout the U.S.

#### The United States as a Product of Gun Violence...
Gun violence has drastically shaped our attitudes and our political climate. The pervasive nature of gun violence in contemporary America is evinced by a map that is entirely constructed of gun violence incident points. 

An interactive map (available online) was constructed to visualize all gun violence incidents. The source code for this map is below: 
```{r eval = FALSE}
gun_v_tidy %>%
	filter(state != "Alaska", state != "Hawaii", longitude < 0) %>% 
	mutate(n_affected_cat = ifelse(n_affected %in% 4:9, 2,
													ifelse(n_affected %in% 10:19, 3,
													ifelse(n_affected %in% 20:34, 4,
													ifelse(n_affected %in% 35:49, 5,
													ifelse(n_affected > 50, 6, 1)))))) %>% 
	mutate(n_affected_cat = as.factor(n_affected_cat),
				 n_affected_cat = recode_factor(n_affected_cat, `1` = "1-3 affected", `2` = "4-9 affected", `3` = "10-19 affected", `4` = "20-34 affected", `5` = "35-49 affected", `6` = "50+ affected"),
				 city_state = str_c(city_or_county, state, sep = ", "), 
				 text_label = str_c(city_state, '\n', year,'\nKilled: ', n_killed, '\nInjured: ', n_injured)) %>% 
  plot_ly(x = ~longitude, y = ~latitude, type = "scatter", mode = "markers",
          alpha = 0.5, 
          color = ~n_affected_cat,
  				colors = "Accent",
  				text = ~text_label) %>% 
	layout(legend = list(x = 0.8, y = 0.1))
```

#### Who Should Care About Gun Violence?

**Anybody!**

+ Gun violence requires people, with population dense states experiencing a greater number of gun violence incidents as seen in the map above.
+ However, even the least populated states still have notable gun violence concerns.
+ The "lethality proportion" is defined as the proportion of those affected by a gun violence incident who are killed in that given gun violence incident; it gets at how deadly a given incident was.
+ Wyoming is the *least densely populated* state in the continental United States, but it hast the *highest lethality proportion* for gun violence incidents. 
+ Other states that follow this low population density ~ high lethality proportion paradigm are Idaho, Montana, Alaska, and Nevada.
+ A possible reason for this paradigm is that lack of population density leads to a lower density of healthcare infrastucture, which then leads to isolation during emergencies. 

The plot below shows the national average for the lethality proportion as a dotted line. 

```{r }
prop_data = 
	gun_v_tidy %>% 
	mutate(prop_killed = n_killed / (n_killed + n_injured))

prop_data %>% 
	summarize(mean = mean(prop_killed))
```

```{r }
prop_data %>% 
	group_by(state) %>% 
	summarize(mean_prop = mean(prop_killed)) %>% 
	mutate(state = forcats::fct_reorder(state, mean_prop)) %>% 
	ggplot(aes(x = state, y = mean_prop)) +
	geom_point() +
	geom_abline(slope = 0, intercept = 0.346, color = "purple", linetype = "dashed") +
	theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
	labs(
    x = " ",
    y = "Lethality Proportion"
  ) 
```

### Section 2: Firearms Mortality (Shiny Dashboard)
Section 2 explores changes in the crude firearm mortality, law strength and unemployment rates across the United States over time. This section is housed in a Shiny Dashboard to enable users to toggle between different variables and years. This section pulls data from the **Dataset 2: CDC Firearm Mortality Data**, **Dataset 3: Gun Law Strength and State Scoring**, and **Dataset 5: Unemployment data** sources. 

The source code for the Shiny dashboard is available to review [online](https://zanis-shiny.shinyapps.io/shiny_states/).  

The three main elements of the Shiny Dashboard are described below:

#### Crude Death Rate Map (2000-2016) for each state
The Crude Death Rate map allows users to toggle between different years (2000-2016), demonstrating the changes in variables over time and visualize regional differences across the U.S. The map includes visualization of distributions of the following variables: 

*  `unemployment rate`, `crude rate`, `deaths`, `population`, `law strength (2016 only)`, `smoking`, `disability`, `self reported health`, `drinking`, `leisure physical activities`, `overweight`, `poverty`, `diabetes`, `mental health`, and `sleep`. 

The map demonstrates that there are significant changes over time in all variables. 

#### Crude Firearm Death Rate Across All States by Age Group and Race
This barplot displays the distribution of firearm mortality deaths by age group and race. Visualizing a breakdown of mortality across different age groups and races is important to identify potential disparities in mortality and identify which groups should be the focus of public health interventions. 

The plot shows that distributions of deaths across age groups and races changes over time. In 2016, the 15-24 age group had the most firearm-related deaths, mostly concentrated in Black/African American group. Over time, white victims comprised a fairly steady number of deaths across all age groups. A comparison of 2000 versus 2016 reveals that deaths among "American Indian or Alaska Natives" have increased across all age groups. Death rates for Asian/Pacific Islanders have consistently been the lowest across all race groups.   

#### Crude Firearm Death Rate Across All States by Age Group and Hispanic Origin
This barplot displays the distribution of firearm mortality deaths by age group and hispanic origin. This was displayed separately than the breakdown by Race because the Hispanic origin variable is assessed separately for this dataset. Moreover, individuals identifying as any race can also choose to select Hispanic/Latino origin. Therefore there is overlap between the two measures. Death certificates missing Hispanic origin information were coded as "not stated" and excluded from this visualization.

Firearm mortality rates are generally higher in Non Hispanic/Latino groups than Hispanic/Latino. In 2016, 

In addition, firearm mortality across all age groups has increased over time.  

### Section 3: Role of Gun Control
Section 3 explores the role of gun control in gun violence, pulling data from the **Dataset 4: Gun Approval Rate and Licensing of States**, **Dataset 3: Gun Law Strength and State Scoring**, and **Dataset 2: CDC Firearm Mortality Data** sources.

#### Approved license vs. law strength
The first plot show the relationship between the percentage of people get approved for gun licenses, mortality rate and the law strength in each state. 
```{r eval = FALSE}
gun_control %>% 
  mutate(text_label = str_c("State:",state,  "\nCrude Rate:", crude_rate)) %>% 
  plot_ly(x = ~approval_rate, y = ~law_strength, 
          type = "scatter", mode = "markers", marker = list(size = ~crude_rate),
          alpha = 0.8, 
          color = ~law_strength,
          text = ~ text_label) %>%
  layout(
    title = "Approval percentage vs. Law Strength",
    xaxis = list(title = "Percentage of people approved for guns among total population"), 
    yaxis = list(title = "Law Strength"),
    annotations = list(
      x = 0.0012,
      y = 55,
      text = "Size of dots shows crude rate",
      xref = "x",
      yref = "y",
      ax = 0,
      ay = 0
))
         
```

As we can see in this plot, in different states, the percentage of people get approved for gun licenses is positively connected with law strength. This is intuitive because it's easier to get approved for guns in states with less strict gun laws. Also, with less strict gun laws, states generally have higher mortality rate. An regression model is built to further explain this relationship in model panel.

#### Application for guns vs. law strength
The second plot shows the relationship application percentage (proportion of people applying for the license in total population), mortality rate and the law strength.

```{r eval = FALSE}
gun_control %>% 
  mutate(text_label = str_c("State:",state, '\nCrude Rate: ', crude_rate)) %>% 
  plot_ly(x = ~application_rate, y = ~law_strength,
          type = "scatter", mode = "markers", marker = list(size = ~crude_rate),
          alpha = 0.8, 
          color = ~law_strength,
          text = ~ text_label) %>%
  layout(
    title = "Application percentage vs. Law Strength",
    xaxis = list(title = "Percentage of people applied for guns among total population"), 
    yaxis = list(title = "Law Strength"),
    annotations = list(
  x = 0.8,
  y = 55,
  text = "Size of dots shows crude rate",
  xref = "x",
  yref = "y",
  ax = 0,
  ay = 0
))

```

As we can see in the plot, 

*  In most states, the percentage of people tried to apply for gun licenses is relatively same across the country, no matter how strict the law is. This shows that people in different state share equal passion to apply for guns.

*  The only exception is Kentucky. According to [Wikipedia](https://en.wikipedia.org/wiki/Gun_laws_in_Kentucky), people don't need to license or permit to own guns for private uses. This explains exceptional passion to apply for guns in Kentucky, leading to an exceptional high proportion of application for background checks. The reason that the total proportion is greater than one might be companies applying for background checks for public gun sales.

### Section 4: Regression Analyses





## Conclusion



